---
title: "Quantum Heart Prediction"
author: "Elvira Khwatenge"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an analysis of a heart disease prediction dataset. The goal is to explore the data, build a predictive model, and evaluate its performance.

## Data Loading and Preprocessing

First, we load the necessary libraries and read the data.

```{r}
library(tidyverse)
library(caret)

data <- read.csv("Heart Prediction Quantum Dataset.csv")
head(data)
str(data)
sum(is.na(data))
```

## Data Processing

preprocess the data by converting categorical variables to factors, checking for outliers, and scaling numerical features.

```{r}
data$Gender <- as.factor(data$Gender)
data$HeartDisease <- as.factor(data$HeartDisease)

summary(data)

boxplot(data$Age, main="Age",col="orange", border="black")
boxplot(data$BloodPressure, main="BloodPressure",col="violet", border="black")
boxplot(data$Cholesterol, main="Cholesterol",col="navy", border="black")
boxplot(data$HeartRate, main="HeartRate",col="red", border="black")
boxplot(data$QuantumPatternFeature, main="QuantumPatternFeature",col="brown", border="black")

Q1 <- quantile(data$Cholesterol, 0.25)
Q3 <- quantile(data$Cholesterol, 0.75)
IQR <- Q3 - Q1
lower_bound <- Q1 - 1.5 * IQR
upper_bound <- Q3 + 1.5 * IQR

data <- data[data$Cholesterol >= lower_bound & data$Cholesterol <= upper_bound, ]

numerical_features <- c("Age", "BloodPressure", "Cholesterol", "HeartRate", "QuantumPatternFeature")
data[numerical_features] <- scale(data[numerical_features])

head(data)
```

###Exploratory Data Analysis (EDA)

perform EDA to understand the relationships between variables.

```{r}
hist(data$Age, main="Age Distribution", xlab="Age", col="skyblue", border="white")
hist(data$BloodPressure, main="Blood Pressure Distribution", xlab="Blood Pressure", col="lightgreen", border="white")
hist(data$Cholesterol, main="Cholesterol Distribution", xlab="Cholesterol", col="lightcoral", border="white")
hist(data$HeartRate, main="Heart Rate Distribution", xlab="Heart Rate", col="lightgoldenrod", border="white")
hist(data$QuantumPatternFeature, main="Quantum Pattern Feature Distribution", xlab="Quantum Pattern Feature", col="lavender", border="white")

# Bar plots with colors
plot(data$Gender, main="Gender Distribution", ylab="Frequency", col=c("pink", "lightblue"))
plot(data$HeartDisease, main="Heart Disease Distribution", ylab="Frequency", col=c("lightgreen", "salmon"))

# Scatter plots with colors
plot(data$Age, data$HeartDisease, main="Age vs Heart Disease", xlab="Age", ylab="Heart Disease", col=ifelse(data$HeartDisease == 1, "red", "blue"), pch=19)
plot(data$BloodPressure, data$HeartDisease, main="Blood Pressure vs Heart Disease", xlab="Blood Pressure", ylab="Heart Disease", col=ifelse(data$HeartDisease == 1, "red", "blue"), pch=19)
plot(data$Cholesterol, data$HeartDisease, main="Cholesterol vs Heart Disease", xlab="Cholesterol", ylab="Heart Disease", col=ifelse(data$HeartDisease == 1, "red", "blue"), pch=19)
plot(data$HeartRate, data$HeartDisease, main="Heart Rate vs Heart Disease", xlab="Heart Rate", ylab="Heart Disease", col=ifelse(data$HeartDisease == 1, "red", "blue"), pch=19)
plot(data$QuantumPatternFeature, data$HeartDisease, main="Quantum Pattern Feature vs Heart Disease", xlab="Quantum Pattern Feature", ylab="Heart Disease", col=ifelse(data$HeartDisease == 1, "red", "blue"), pch=19)

# Select numeric columns and handle missing values
numerical_features <- c("Age", "BloodPressure", "Cholesterol", "HeartRate", "QuantumPatternFeature")
numeric_data <- data[, numerical_features]
numeric_data <- na.omit(numeric_data)

# Remove columns with zero variance (if any)
numeric_data <- numeric_data[, apply(numeric_data, 2, var) != 0]

# Compute correlation matrix
correlation_matrix <- cor(numeric_data, use = "complete.obs")

# Visualize with corrplot
library(corrplot)
corrplot(correlation_matrix, method = "color", type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)



```


###  Model Building 
Build a logistic regression model to predict heart disease


```{r}
set.seed(123)
train_index <- createDataPartition(data$HeartDisease, p = 0.7, list = FALSE)
train_data <- data[train_index, ]
test_data <- data[-train_index, ]

# Train the logistic regression model
model <- glm(HeartDisease ~ ., data = train_data, family = "binomial")

# Summarize the model
summary(model)

# Make predictions on the test data
predictions <- predict(model, newdata = test_data, type = "response")
predicted_classes <- ifelse(predictions > 0.5, 1, 0)

# Evaluate the model
confusionMatrix(as.factor(predicted_classes), test_data$HeartDisease)
```


# Discussion and Interpretation

## Data Summary and Preprocessing
At the Initial Data Inspection the dataset consisted of 500 observations with 7 variables, including Age, Gender, BloodPressure, Cholesterol, HeartRate, QuantumPatternFeature, and HeartDisease. No missing values were detected.

During Data Type Conversion the Gender and HeartDisease variables were converted to factor variables for appropriate analysis.

Outliers in the Cholesterol variable were removed to ensure data quality.

Numerical features were scaled to standardize their ranges for model performance.

## Exploratory Data Analysis (EDA)
### Histograms

The age histogram shows a relatively normal distribution of ages, with a slight right skew, indicating a higher proportion of middle-aged and older individuals.

The bimodal blood pressure distribution suggests two distinct groups, possibly representing normal and high blood pressure populations.

The right-skewed distribution indicates that most individuals have normal cholesterol levels, but a notable subset has higher levels.

The normal distribution indicates that most individuals have heart rates within the typical range.

Quantum Pattern Feature/multimodal Distribution suggests distinct subgroups within the population relevant to heart disease prediction.

### Bar Plots

The distribution of males and females in the dataset is critical for understanding potential gender-based risk differences.

Heart Disease Distribution Illustrates the prevalence of heart disease in the sample, providing insight into the overall burden of the disease.

Scatter plots of Age, BloodPressure, Cholesterol, HeartRate, and QuantumPatternFeature versus HeartDisease use color-coding to visualize relationships. Patterns or clusters reveal risk factors or predictive indicators.

Correlation Matrix- heatmap visualizes correlations between numerical features, with stronger correlations indicating potential multicollinearity, which affects model interpretation and feature selection.

### Model Building- Logistic Regression

Coefficients

1. Intercept: 1.7752

2. Age: -0.2477 (p = 0.368)

3. Gender1: 0.2093 (p = 0.676)

4. BloodPressure: -0.1250 (p = 0.595)

5. Cholesterol: 0.1128 (p = 0.679)

6. HeartRate: -0.2070 (p = 0.397)

7. QuantumPatternFeature: -7.6702 (p < 0.001)

### Interpretation:

The QuantumPatternFeature is highly significant (p < 0.001), indicating a strong relationship with heart disease.

Other variables such as Age, Gender, BloodPressure, Cholesterol, and HeartRate are not statistically significant in this model (p > 0.05).

### Model Evaluation (Confusion Matrix)

Accuracy: 93.33%

Sensitivity: 96.67%

Specificity: 91.11%


# Results Interpretation

## Demographic and Health Indicators

A slightly right-skewed distribution suggests more middle-aged and older individuals, which aligns with typical heart disease demographics. Blood Pressure/Bimodal distribution could indicate a mix of individuals with normal and high blood pressure, requiring further stratified analysis.Right-skewness suggests a need for targeted cholesterol management strategies. Normal distribution suggests heart rates are generally within expected ranges. Quantum Pattern Feature/ Multimodal distribution suggests subgroups requiring further investigation to understand patterns related to heart disease.

#### Gender and Heart Disease Distribution

Knowing the proportions of males and females helps understand potential gender-specific risk factors. Heart Disease Distribution Indicates overall disease burden, useful for resource allocation and public health planning.

#### Relationships Between Variables and Heart Disease

Color-coding aids in visualizing relationships between variables and heart disease in correlation matrix, highlights potential risk factors.Correlation Matrix helps identify multicollinearity, informing model adjustments for more accurate predictions.

## Model Performance
The logistic regression model shows high accuracy (93.33%), sensitivity (96.67%), and specificity (91.11%), demonstrating strong predictive capability. The highly significant QuantumPatternFeature suggests it is a crucial predictor of heart disease within this model.

## Implications
For Academia- Validate QuantumPatternFeature in other datasets, explore interactions, and test other models like Random Forests or Neural Networks.

For Industry- Develop diagnostic tools incorporating the QuantumPatternFeature, and create personalized health monitoring devices.

For Policy Makers- Implement targeted interventions for high-risk groups identified by the model, focusing on QuantumPatternFeature and traditional risk factors.

## Caveats
The model is based on a specific dataset and may not generalize to other populations.

Some features might be correlated, affecting coefficient interpretation.Some variables lack statistical significance (p > 0.05), requiring cautious interpretation.
